File-based localization of numerical perturbations in data analysis pipelines

### 2020-04-03 paper.tex
add figure script and update some comments

\toolname identifies the processes in a pipeline that produce different
results when executed in two different conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified (Step 2) and
disambiguated, resulting in a provenance \emph{DAG} (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Step 3)
\tristan{Ali, this is not true in Figure 1: at step 3 there are still
transient files}. DAGs produced in the two conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Step
4). Finally, a labeled graph is produced that identifies the
non-reproducible processes in the pipeline.

@@

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot, for instance
when complete different processing paths are used in different subjects. \toolname 
includes helpers to identify different instances of provenance graphs \tristan{mention tree distance here}.

@@

\begin{figure}
\begin{subfigure}{0.4\linewidth}
  [ FIGURE GOES HERE ]

  \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
  \label{fig:provenance-graph}
\end{subfigure}
\hfill
\begin{subfigure}{0.4\linewidth}
  [ FIGURE GOES HERE ]

  \caption{Provenance DAG, with disambiguated transient files.}
  \label{fig:provenance-dag}
\end{subfigure}
  % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  created from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles, files with rectangles, and read/write accesses 
  with plain edges. 
  For visualization purposes, the process tree is represented too, with
  dashed edges. Every node in the graph is labeled using (1) a process id
  created by our reconstruction, (2) the name of the executable run by the
  process or the file name. Process 0 is the initial call to the
  \texttt{sh} interpreter; the other processes are the calls to FSL bet,
  stats and maths made in Listing~\ref{listing:sample-script}. Process 73
  is forked by process 2: it was captured by \reprozip while it did not
  appear in Listing~\ref{listing:sample-script}.
}
  \label{fig:simple_script}
\end{figure}

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
on PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done by these
pipelines is available in~\cite{glasser2013}. 
In summary, PreFreeSurfer consists of the following steps: 
\begin{itemize}
\item Distortion Correction (DC), 
\item Anatomical Average (AAve), 
\item Anterior/Posterior Commissure Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And FreeSurfer consists of the following ones: \tristan{Check these steps}
\begin{itemize}
\item Image downsampling, 
\item T1w image registration, 
\item Surface placement, 
\item Surface registration.
\end{itemize}
The average processing time per subject is approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
For each subject, available data consisted of 1 or 2 T1 images, and 1 or 2 T2 images. 
% \tristan{Can you let me know where the data is located, so that I could describe it here?
% Subjects are located in consider:/data/asalari/ali-tests/PFS\_Centos6\_Traced
% and consider:/data/asalari/ali-tests/PFS\_Centos7\_Traced 
% and outputs are in consider:/data/asalari/ali-tests/nurm-out}
Acquisition protocols and parameters are detailed in~\cite{van2013wu}. \tristan{Add summary of matrix sizes.}

@@

Figure~\ref{fig:pfs_freq} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average). \tristan{figure and caption to be fixed.}

\tristan{Ali, it would be useful to show a complete PreFreesurfer graph here.}

% Between subject differences can be a result of different subject types. 
% For instance, the subjects contain two T1 images need more calculations 
% to produce an average image between them.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  \tristan{Remove frequency values as they are shown by the color already. Remove multipliers, multiply cells instead. Cells should just contain subject numbers.}
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the FLIRT tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 95\% of the subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  85\% of the subjects.
        }
  \label{fig:pfs_freq}
\end{figure*}

To visually illustrate the observed differences,
Figure~\ref{fig:fnirt_result} compares FNIRT results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes. 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between T2 FNIRT results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7) \tristan{Improve contrast}. An animated version of the comparison is available 
    \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {here}.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

The only non-reproducible process identified by \toolname in FreeSurfer was
\texttt{mris\_make\_surfaces} (cortical and white matter surfaces
generation), which is also the only dynamically-linked executable in this pipeline
\tristan{Ali, could you check that this is indeed the only dynamic
executable in the pipeline?}. \texttt{mris\_make\_surfaces} produced different results for
10 out of 20 subjects.

@@

\begin{figure*}
  \centering
  \begin{subfigure}{0.45\linewidth}
    [FIGURE GOES HERE]
    \caption{Regions with \emph{less} than 10,000 voxels.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\linewidth}
    [FIGURE GOES HERE]
    \caption{Regions with \emph{more} than 10,000 voxels.}
  \end{subfigure}
    % \includegraphics[width=\textwidth]{images/scatter_plot.png} 
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).
        To improve readability, the two graphs are on different y scales. \tristan{Ali, could you separate the figures in 2 and insert them in each subfigure?}} 
    \label{fig:scatter_plot}
  \end{figure*}

@@

The processes identified by \toolname as non-reproducible were all
associated with dynamically-linked executables \tristan{Ali, please check
that}. This makes complete sense as statically-linked executables wouldn't
be impacted by library updates. Moreover, the hypothetical effects of
hardware or Linux kernel updates were not measured, as the different
operating sytems were deployed in Docker containers on the same host, that
is, using the same kernel and hardware.

>>>>

\toolname identifies the processes in a pipeline that produce different
results when executed in two different conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified (Step 2) and
disambiguated, resulting in a provenance \emph{DAG} (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Step 3). 
DAGs produced in the two conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Step
4). Finally, a labeled graph is produced that identifies the
non-reproducible processes in the pipeline.

@@

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot, for instance
when complete different processing paths are used in different subjects. \toolname 
includes helpers to identify different instances of provenance graphs.
For this purpose, \toolname measures the distance between process trees using 
the tree edit distance method~\cite{zhang1989simple}, as implemented in 
the zss Python package. This distance is defined as the minimum number of
edit operations to transform one tree into the other. Three edit operations
are considered: node label modification, node removal, and node insertion.
Each operation has an associated cost of 1.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph}
  \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
  \label{fig:provenance-graph}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph-dag}
  \caption{Provenance DAG, with disambiguated transient files.}
  \label{fig:provenance-dag}
\end{subfigure}
  % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  created from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles, files with rectangles, and read/write accesses 
  with plain edges. 
  For visualization purposes, the process tree is represented too, with
  dashed edges. Every node in the graph is labeled using (1) a process id
  created by our reconstruction, (2) the name of the executable run by the
  process or the file name. Process 0 is the initial call to the
  \texttt{sh} interpreter; the other processes are the calls to FSL bet,
  stats and maths made in Listing~\ref{listing:sample-script}. Process 73
  is forked by process 2: it was captured by \reprozip while it did not
  appear in Listing~\ref{listing:sample-script}.
}
  \label{fig:simple_script}
\end{figure}

@@

The HCP developed a set of pre-processing pipelines to process structural,
functional, and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, and particularly
on PreFreeSurfer and FreeSurfer. 
A detailed description of the analyses done by these
pipelines is available in~\cite{glasser2013}. 
In summary, PreFreeSurfer consists of the following steps: 
\begin{itemize}
\item Distortion Correction (DC), 
\item Anatomical Average (AAve), 
\item Anterior/Posterior Commissure Alignment (ACPC-A), 
\item Brain Extraction (BExt), 
\item Bias Field Correction (BFC), 
\item Atlas-Registration (AR).
\end{itemize}
And FreeSurfer consists of the following ones:
\begin{itemize}
\item Image downsampling, 
\item T1w image registration, 
\item T1w image segmentation,
\item Surface placement, 
\item Surface registration.
\end{itemize}
The average processing time per subject is approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size is
approximately 2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}

We randomly selected 20 unprocessed subjects 
from the HCP data release S500 
available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
For each subject, available data consisted of 1 or 2 T1 images, and 1 or 2 T2 images. 
The voxel size of the acquired T1 and T2 images was 0.7 × 0.7 × 0.7 mm, 
and the dimension of the images was 256 x 320 x 320.
% \tristan{Can you let me know where the data is located, so that I could describe it here?
% Subjects are located in consider:/data/asalari/ali-tests/PFS\_Centos6\_Traced
% and consider:/data/asalari/ali-tests/PFS\_Centos7\_Traced 
% and outputs are in consider:/data/asalari/ali-tests/nurm-out}
Acquisition protocols and parameters are detailed in~\cite{van2013wu}.

@@

Figure~\ref{fig:pfs_freq} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
Differences were observed in linear registration 
with FSL FLIRT (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL FNIRT (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL maths  (in Anatomical Average). 
A complete provenance graph of the PreFreeSurfer is illustrated in Figure~\ref{fig:complete_pfs}
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/pfs-intransient} 
    \caption{A complete provenance graph from the PreFreesurfer pipeline.
     Full-resolution image is available 
     \href{https://drive.google.com/file/d/1xonnMq4hLEe6GwMp4P7umikvrIAeflDe/view?usp=sharing}{here}.}
} 
    \label{fig:complete_pfs}
\end{figure}
% Between subject differences can be a result of different subject types. 
% For instance, the subjects contain two T1 images need more calculations 
% to produce an average image between them.

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the FLIRT tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
        }
  \label{fig:pfs_freq}
\end{figure*}

To visually illustrate the observed differences,
Figure~\ref{fig:fnirt_result} compares FNIRT results in Brain Extraction
for a particular subject. Differences appear to be very substantial, in
particular in the frontal and parietal lobes. 
\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/t2w_alignment.png} 
    \caption{Differences between T2 FNIRT results in PreFreeSurfer's Brain Extraction (CentOS6 vs 
    CentOS7). An animated version of the comparison is available 
    \href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/images/pfs_t2w_alignment.gif}
    {here}.
} 
    \label{fig:fnirt_result}
\end{figure}

@@

The only non-reproducible process identified by \toolname in FreeSurfer was
\texttt{mris\_make\_surfaces} (cortical and white matter surfaces
generation), which is also the only dynamically-linked executable in this pipeline
\tristan{Ali, could you check that this is indeed the only dynamic
executable in the pipeline?
Ali: There is another dynamically-linked executable named mris\_curvature 
that doesn't create difference when set a seed for that.
Also, another executables are in directory freesurfer/mni/bin/ such as volume\_hist, 
mincinfo, mincresample, \dots
that are dynamically-linked but do not create differences.}. 
\texttt{mris\_make\_surfaces} produced different results for
10 out of 20 subjects.

@@

\begin{figure*}
  \centering
  \begin{subfigure}{0.45\linewidth}
    \includegraphics[width=\linewidth]{images/scatter_plot_1.png}
    \caption{Regions with \emph{less} than 10,000 voxels.}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.45\linewidth}
    \includegraphics[width=\linewidth]{images/scatter_plot_2.png}
    \caption{Regions with \emph{more} than 10,000 voxels.}
  \end{subfigure}
    % \includegraphics[width=\textwidth]{images/scatter_plot.png} 
    \caption{Dice coefficients between regions segmented by FreeSurfer in CentOS6 vs CentOS7 (N=20).
        To improve readability, the two graphs are on different y scales.}
    \label{fig:scatter_plot}
  \end{figure*}

@@

The processes identified by \toolname as non-reproducible were all
associated with dynamically-linked executables. 
This makes complete sense as statically-linked executables wouldn't
be impacted by library updates. Moreover, the hypothetical effects of
hardware or Linux kernel updates were not measured, as the different
operating sytems were deployed in Docker containers on the same host, that
is, using the same kernel and hardware.
